Next Trippy Structures\thanks{Supported by organization x.}

### 2019-03-08 paper.tex
Daniil review

If possible, cite applications of our kind of analysis for transportat/ad companies, to show that what we are doing is useful. One example is \cite{tu2018spatial}. Another is \cite{zhang2017targeted}. And by the way, the guys in \cite{weng2018mining} do some analysis with data that has already boarding and alighting stops!

@@

\subsubsection{Network constrained trajectory indexing}
There are also numerous approaches that use R-Tree based indexes for trajectories that are constrained to an underlying network, aiming to decrease the redundancy in the representation by separating the representation into levels. Examples of such indexes include the {\em FNR-Tree}~\cite{DBLP:conf/ssd/Frentzos03}, where the network elements are indexed with a 2D R-Tree and 1D R-Trees to index time intervals of moving objects along the edges of the network. Some of the limitations of the FNR-Tree are addressed in \cite{DBLP:journals/geoinformatica/AlmeidaG05}, where the authors propose the {\em MON-Tree}, where the moving objects are indexes in two dimensional R-Trees (the dimensions being edge position vs time). More recent alternatives also integrate {\em B$^+$-Trees}, which has proven to be more space and time efficient for indexing the temporal dimension as proven by the {\em TMN-Tree}~\cite{chang2010tmn}. Alternatively, in \cite{rivera2018faster} a compact representation of time intervals is proposed using two bitvectors, that can be combined with these R-Tree based indexes to increase the efficiency for temporal filtering. Refer to \cite{john2017performance} for a quick comparison of these R-Tree based indexes.

As a competitive alternative to these R-Tree based indexes, {\em PARINET}~\cite{DBLP:journals/vldb/PopaZOBV11}, that builds spatial partitions from the trajectories based on their distribution and network topology, and uses a B$^+$-Tree to index the trajectories in each partition by time intervals. Although candidate trajectories must be filtered in memory during queries, PARINET is highly efficient in practice as it minimizes the number of disk accesses needed.
%The same ideas were used in {\em TRIFL}~\cite{that2015trifl}, where the cost model is adapted for flash storage.

@@

A recent compact data structure names {\em CiNCT} has been proposed in \cite{koide2018cinct}, where trajectories are encoded into a string, that is used to build an FM-index \cite{DBLP:conf/focs/FerraginaM00} with a Huffman-shaped Wavelet Tree \cite{ferragina2009compressed}. To further save space, the string is constructed with relative movement labels instead of absolute edges, with an auxiliary structure that represents a network graph built from the input trajectories themselves. While this structure excels at pattern-matching and path extraction in vehicular networks (such as the streets of a city), it cannot be applied to our problems in public transportation networks.

\subsection{Trajectory data mining}
Currently there are several known techniques that would allow to collect data regarding users' trips over a public transportation network. Numerous works exist where these trajectories are mined from the transactions of smart cards \cite{bhaskar2015passenger,wang2014aggregated}. This can be complemented with information derived from GPS devices, as shown in \cite{ma2014development}. Alternatively, these trajectories may be extracted relying on positioning obtained from cellular networks, as proven by \cite{liu2017exploring}.

In addition, the authors of \cite{tao2014exploring} have specifically tackled the challenge of reconstructing full trajectories, accounting for trip-chaining, using data obtained from smart cards. Furthermore, in \cite{alsger2016validating} it is also proven that not only the alighting stops, but also the (last) destination stop of a trip can be estimated from boarding data gathered by a smart card, within a reasonable accuracy.

This summarized review proves that, although we were not able to access real data from a public transportation company, such curated information about users' trips can indeed be obtained and used for our proposed representation.

@@

\subsubsection{Compressed Suffix Array (CSA)}
\label{sec:csa}
One common structure to count or locate all the occurrences of a pattern $P[1,m]$ in a string $S[1,n]$ of symbols from an alphabet $\Sigma$ is the Suffix Array (SA) \cite{MM93}. The SA lexicographically sorts all the suffix of $S$ and builds an array $A[1,n]$ with their indexes, holding that $S[A[i],n] \prec S[A[i+1],n]$. This allows to find out all the occurrences of a substring $P[1,m]$ in $O(m \log n)$ with binary searches.

A simple enhancement to avoid storing the original string $S$ is to align $A$ with a bitvector $D[1,n]$ where $D[i]=1~\Leftrightarrow~S[A[i-1]] \prec S[A[i]]$. That is, $D$ marks with a 1 all the indexes $i$ of $A$ where the starting symbol of that suffix $S[A[i]]$ has a larger lexicographical value than the one from the previous suffix $S[A[i-1]]$. With $D$, keeping $S$ is not longer needed as $S[A[i]] = rank_1(D,i)$, where $rank_1(D,i)$ returns the number of ones in $D[1..i]$.

@@

\subsubsection{Wavelet Matrix (WM)}
\label{sec:wm}
Based on the Wavelet Trees \cite{WT03}, the Wavelet Matrix \cite{CNO15} is a structure that is able to perform, among others, operations of $rank_a(S,i)$ and $select_a(S,i)$ $\forall a \in \Sigma$ and a sequence $S[1..n]$ over the vocabulary $\Sigma$ in $O(log\sigma)$ time while using $n\lceil log_2\sigma\rceil~+~o(n)$ bits of space, begin $\sigma = |\Sigma|$.

The WM is built with $\lceil log_2\sigma\rceil$ levels, each level having a bitvector $B_i$ with rank and select capabilities, where $B_1$ contains the most significant bit of every symbol $a \in S$, $B_2$ contains the second most significant bit and so forth. To further explain how these bits are positioned, we must consider $a_k$ as the $k$th bit of $a$ in significance, and $B^{-1}_i[a_k]$ as the position of the bit $a_k$ from $a$ in the bitvector $B_i$. Then, 
\[
    B_i[j] = \left\{\begin{array}{lr}
        S[j]_1, & \text{iff } i=1\\
        a_i, & \text{iff } i>1 \wedge (a_{i-1}=0 \wedge j=rank_0(B_{i-1},B^{-1}_{i-1}(a_{i-1}))) \\
        & \vee~(a_{i-1}=1~\wedge~j=rank_0(B_{i-1},n)~+~rank_1(B_{i-1},B^{-1}_{i-1}(a_{i-1})))
        \end{array}\right.
\]

@@

This structure allows to answer the following operations in $O(log\sigma)$ time:
\begin{itemize}
    \item $access(S,i)$ returns the value in any position in the original sequence $S$ by following through the levels of the WM using the $rank$ operation on bitvectors. If $B_1[i]$ is 0, then $B_2[rank_0(B_1,i)]$ must be accessed next. Otherwise, it is $B_2[rank_0(B_1,n) + rank_1(B_1,i)]$. This must be recursively applied until the leaf level is reached, and the traversed bits will compose the value of $S[i]$.
    \item $rank_a(S,i)$ is calculated with the same operations as in $access$, with the difference that in this case we know whether we need to use $rank_0$ or $rank_1$ on each bitvector as the value of the symbol $a$ is known. To calculate the number of occurrences of $a$ in $S[1..i]$ we must simply subtract the position reached in the leaf level to the position of the first occurrence of $a$ on that level, which can be either stored for each symbol in $O(\sigma)$ bits or calculated by making an analogous traversal from the first position of $S$.
    \item $select_a(S,i)$ first $i$ must be added to the first position of $a$ in the leaf level, which as for $rank$, can be either stored or calculated. After obtaining this position $j$, the WM must be traversed in reverse order using $select$ on bitvectors until the first level is reached. When $a_i=0$ the traversal from the lower level position $B_{i+1}[j]$ must be as $select_0(B_i,j)$, while when $a_i=1$, it is done as $select_1(B_i,j-rank_0(B_{i},n))$.
    \item $range_{a,b}(S,i,j)$ returns the number of occurrences of the symbols between $a$ and $b$ in $S[i..j]$. This operation was first introduced in \cite{gagie2012new} for Wavelet Trees and later supported by the Wavelet Matrix \cite{CNO15}, and we make extensive use of it in this work. It is solved with two simultaneous traversals similar to $rank$ over $i$ and $j$, only that it is possible that at some levels it is required to calculate both $rank_0$ and $rank_1$. However, the worst case time complexity of $O(log\sigma)$ is maintained, due to the fact that it is not necessary to reach the leaf level for all the symbols in $a..b$\footnote{As long as $a\neq b$. Otherwise, $rank_a(S,j)-rank_a(S,i)$ can be used.}.
\end{itemize}

@@

\section{Our model for compact trip representation}
In our previous structures proposed in \cite{brisaboa2018new}, we propose a work-in-progress model to represent the demand information for a public transportation network, as seen in the Figure~\ref{fig:er}.

@@

These are the main elements from our model:
\begin{itemize}
    \item A \textbf{stop\_place} is a physical stop with a location, on which several lines may make stops.
    \item A \textbf{line} is an ordered sequence of stop places that can be traveled by a transport vehicle, such as a bus or a train. It only considers one direction. For this reason, there is often a different and complementary line for the opposite direction, as shown in the Figure~\ref{fig:example_network}.
    \item A \textbf{journey} is a single traversal of a transport vehicle over a line. It can be seen as a vehicle trip, instead of a user trip.
    \item A \textbf{stage} is formed by a boarding from a stop and alighting to another from the same single line and journey.
    \item An \textbf{user\_trip} is a concatenation of several stages, until the final destination (alighting stop of the last stage) is reached.
\end{itemize}

@@

This approach allows us to treat the information in a layered fashion: the bottom layer is a static network representation, formed by the line and stop\_place entities, the middle layer are the journeys made by vehicles that make stops at specific times, while the top layer are the trips made by the users over these vehicle journeys. In the Figure~\ref{fig:example_network} we show an example for the two bottom layers. Our proposed implementation can be found in Section~\ref{sec:ps}.

@@

\section{Proposed Structures}
\label{sec:ps}
As of today, we are not aware of any indexing structure that would allow to efficiently represent trajectories that could also support all the kinds of queries described in Section~\ref{sec:rq}. For this reason, we propose two data structures, \acumm~and~\ctr. The former is targeted for queries of type A, solving most aggregation queries in constant time, while the later can be used for queries of type B.

\subsection{Common Structures}
\label{sec:cs}
Both \acumm~and \ctr~need a network representation to be operated, along with information about the transport vehicle journeys. Therefore, the following structures represent a network with stops $s_i \in S = \{1,...|S|\}$, lines $l_i \in L = \{1,...|L|\}$ and journeys $j_i \in J^l = \{1,...|J^l|\}$. It is important to state that journeys are \textbf{not} identified by $j_i$, as the same $j_i$ can belong to several $J^l$ from different lines, so we speak about journey \textbf{codes} (jcodes) instead of journey identifiers.

@@

\ttctr was introduced in \cite{brisaboa2018new}, where the spatial component (the pairs $(s,l)$ for the stops and lines of a trip) is represented with a \texttt{CSA} where each unique pair $(s,l)$ is encoded as an integer $id$ in the input sequence $T[1,n]$ that is used to build the \texttt{CSA}.

@@

\begin{figure}[hbt!]
\includegraphics[width=0.4\textwidth]{example_trips.eps}
\caption{An example five trips sorted as required by \ttctr and \ctr}
\label{fig:example_trips}
\end{figure}

We also need an unambiguous function to encode the pairs $(s,l)$. Consider a vocabulary $V$ such that:
\begin{itemize}
	\item Entry $V[0]$ is reserved for the terminator symbol $\$$.
	\item Entries $\langle V[1],V[2], \dots V[|S|]\rangle]$ are associated to stops $s_1,s_2,\dots, s_{|S|}$ and are used to represent the final stops of the trips. That is, when a given stop $s_i$ ends a user trip, it is given $id \leftarrow s_i$.
	\item The following $|L|$$\times$$|S|$ entries are associated to the sequence composed of the pairs $(s,l) \in S\times L$, sorted first by the stop id $s$ and later by the line id $l$. That is, entry $V[|S|+1]$ is given to $(s_1,l_1)$; $V[|S|+2]$ to $(s_1,l_2)$; $V[|S|+3]$ to $(s_1,l_3)$; $\dots$; $V[|S|+|L|]$ to $(s_1,l_{|L|})$;  $V[|S|+|L|+1]$ to $(s_2, l_1)$, $V[|S|+|L|+2]$ to $(s_2, l_2)$, and so on. Therefore, it is easy to see that any $(s_i,l_j)$ is going to be associated to the entry $V[|S|+ |L|(i-1) + j]$.
\end{itemize}

While this arrangement would theoretically produce many entries in $V$ that are mapped to pairs $(s,l)$ that are unused in $T$, either because the stop is never traversed by that line or because we do not have the record of a user trip containing it, these entries can be skipped with a compact bitvector $B$ with rank and select capabilities, that marks with a one every used entry from $V$. This will enable us to operate with a much smaller vocabulary $V'$ with only the used entries from $V$, such that $V[i] = V'[rank_1(B,i)]$. Refer to the vocabulary shown Figure~\ref{fig:ttctr}(2) for an example where pairs $s:l$ are encoded to 43 unique identifiers in $V$. After that, $B$ marks which of the entries of $V$ actually appear in the original sequence. Finally $V'$ is built with only 12 entries, for each one from $B$. Note that neither $V$ nor $V'$ are explicitly represented in practice, as $rank$ and $select$ operations over $B$ are enough to map and unmap, respectively, vocabulary identifiers.

@@

After this, the sequence $T[1,n]$ is built, with the identifiers obtained from mapping mapping in the vocabulary entries of $V'$, over which a \texttt{CSA} is built, as seen in Figure~\ref{fig:ttctr}(3). Each encoded trip in $T$ is terminated with with additional $\$$ symbols. While in the final \texttt{CSA} we assign all these $\$$ a lexicographical value of 0 $(V[0])$, we assign them different values during the construction of the suffix array (A) to ensure that the entries for $\$$ in $A$ maintain the same order as in the original text. Finally, we make a modification on $\Psi$ to make the entries of each $\$$ point to the start of its own trip instead of the next one. These two modifications are proven necessary for our implemented queries, at the expense of losing some of the properties of a classic CSA. For reference, in Figure~\ref{fig:ttctr} we also $A'$ and $\Psi'$, that show how our modifications compare to the original CSA.

@@

With \ttctr~we obtain a clear separation between the spatial representation of the trips (\texttt{CSA}) and the temporal representation (\texttt{WM} of $jcodes$), where the former can be used to address queries such as ``number of passengers that started their trip from a stop $s\in S$ and a line $l\in L$'' (\texttt{start\_X$_{L}$} from Section~\ref{sec:rq}) with a binary search of the pattern $\$,s:l$, while the later can be used to filter down these results to a time window (\texttt{start\_X$_{LT}$}) with a $range_{a,b}(S,i,j)$ operation over the \texttt{WM}, where $a$ and $b$ are $jcodes$ obtained from Algorithm~\ref{alg:jcodes} and $i$ and $j$ delimit the range of the results obtained in $\Psi$. Because the $\$$ symbols were made cyclical in $\Psi$, it is also possible to answer \texttt{from\_X\_to\_Y} queries by searching for a pattern $Y,\$,X:l$ instead.

\subsection{\ctr}
\label{sec:ctr}
A fundamental weakness of \ttctr~is that it requires several binary search operations over the \texttt{CSA} in the following cases:
\begin{itemize}
    \item We are interested in the number of passengers that started their trip at a stop $s$ and a time window $t_a...t_z$, but from \textbf{any line} (\texttt{start\_X$_{T}$} from Section~\ref{sec:rq}). As $jcodes$ are relative to lines, we must make a separate query for each possible $(s:l_i) \forall l_i \in L$ combination.
    \item We need to restrict the line of a final stop, in queries such as \texttt{end\_X$_{L}$} or \texttt{from\_X\_to\_Y$_{L}$} (and similar variations). Because the final stops belong to separate entries of the vocabulary that do not encode line identifiers, to restrict a stop $Y\in S$ to a line $l\in L$ we need to search for every possible expanded pattern $W:l,Y...$, for every stop $W$ from the line $l$ that could have been boarded before alighting at $Y$. While it looks tempting to address this issue by modifying the design of \ttctr~so that final stops also encode line identifiers, this would in turn make queries that do not restrict the line of the final stop inefficient, and we would need to perform a new query for every combination of $(s:l_i) \forall l_i \in L$, as in the previous case. \marginpar{\tiny Viendo los experimentos, uno se puede preguntar si no habria quedado mejor meter lineas en paradas finales...}
\end{itemize}

These weakness motivated us to develop a second version, \ctr, which reduces the complexity of the queries that restrict the final line, and delegates line checks on a new WM, allowing for more space-time trade-off. As in \ttctr, the input trips need to be sorted by the same criteria, but in \ctr~we use three complementary structures to represent each component of the sequence, as shown in Figure~\ref{fig:example_xctr}:
\begin{enumerate}[(i)]
    \item An adapted Compressed Suffix Array (\texttt{CSA}) over the stop identifiers of all trips, concatenated into a string with additional terminator symbols $\$$ appended at the end of each trip. As in the \texttt{CSA} from \ttctr, we make these $\$$ symbols maintain the order of the trips and cyclical in $\Psi$. Because this time we do not encode line and stop identifiers together and \texttt{CSA} only encodes stops, there is no need for a complex vocabulary anymore.
    \item \texttt{WML}: Aligned to the entries of (i) there is a Wavelet Matrix (WM) for the line identifiers of each stop. Aligned to the $\$$ section we duplicate the starting lines of each trip. As a trivial optimization, we build a separate WM for every stop, allowing us to save space due to the fact that a single stop does not usually belong to many lines, thus the average height of these WM is no larger (and usually smaller) than the height of a single WM.
    \item \texttt{WMJ}: A WM of jcodes aligned to the leaves of (ii). Note that this makes this structure dependant on (ii), which is coherent with the fact that journey codes themselves are relative to the line identifier. In case (ii) implements the optimization described, the entries of the WM must also be rearranged to match the delimited stops.
\end{enumerate}

@@

    \item In \textbf{line 16} we return the number of entries of \texttt{WMJ} between \DataSty{j$_a$} and \DataSty{j$_z$} within our final subrange. The function \FuncSty{GetCount} does not report any range boundaries as \FuncSty{GetRange} does, but simply returns the number of occurrences. Of course, the time complexity of this operation is still proportional to the height of the WM even though the matching entries can be scattered (not delimiting a contiguous range).
\end{itemize}

>>>>

If possible, cite applications of our kind of analysis for transport/advertising companies, to show that what we are doing is useful. One example is \cite{tu2018spatial}. Another is \cite{zhang2017targeted}. And by the way, the guys in \cite{weng2018mining} do some analysis with data that has already boarding and alighting stops!

@@

\subsubsection{Network constrained trajectory indexing}
There are also numerous approaches that use R-Tree based indexes for trajectories that are constrained to an underlying network, aiming to decrease the redundancy in the representation by separating the representation into levels. Examples of such indexes include the {\em FNR-Tree}~\cite{DBLP:conf/ssd/Frentzos03}, where the network elements are indexed with a 2D R-Tree and 1D R-Trees to index time intervals of moving objects along the edges of the network. Some of the limitations of the FNR-Tree are addressed in \cite{DBLP:journals/geoinformatica/AlmeidaG05}, where the authors propose the {\em MON-Tree}, where the moving objects are indexes in two dimensional R-Trees (the dimensions being edge position vs time). More recent alternatives, such as the {\em TMN-Tree}~\cite{chang2010tmn}, integrate {\em B$^+$-Trees}, which have proven to be more space and time efficient for indexing the temporal dimension. Alternatively, in \cite{rivera2018faster} a compact representation of time intervals is proposed using two bitvectors, that can be combined with these R-Tree based indexes to increase the efficiency for temporal filtering. Refer to \cite{john2017performance} for a quick comparison of these R-Tree based indexes.

As a competitive alternative to these R-Tree based indexes, {\em PARINET}~\cite{DBLP:journals/vldb/PopaZOBV11}, builds spatial partitions from the trajectories based on their distribution and network topology, and uses a B$^+$-Tree to index the trajectories in each partition by time intervals. Although candidate trajectories must be filtered in memory during queries, PARINET is highly efficient in practice as its partitioning strategy minimizes the number of disk accesses needed.
%The same ideas were used in {\em TRIFL}~\cite{that2015trifl}, where the cost model is adapted for flash storage.

@@

A recent compact data structure named {\em CiNCT} has been proposed in \cite{koide2018cinct}, where trajectories are encoded into a string, that is used to build an FM-index \cite{DBLP:conf/focs/FerraginaM00} with a Huffman-shaped Wavelet Tree \cite{ferragina2009compressed}. To further save space, the string is constructed with relative movement labels instead of absolute edges, with an auxiliary structure that represents a network graph built from the input trajectories themselves. While this structure excels at pattern-matching and path extraction in vehicular networks (such as the streets of a city), it cannot be applied to our problems in public transportation networks.

\subsection{Trajectory data mining}
Currently there are several known techniques that would allow to collect data regarding users' trips over a public transportation network. Numerous works exist where those trajectories are mined from the transactions of smart cards \cite{bhaskar2015passenger,wang2014aggregated}. This can be complemented with information derived from GPS devices, as shown in \cite{ma2014development}. Alternatively, reliable trajectories may be extracted relying on positioning obtained from cellular networks, as proven by \cite{liu2017exploring}.

In addition, the authors of \cite{tao2014exploring} have specifically tackled the challenge of reconstructing full trajectories, accounting for trip-chaining, by using data obtained from smart cards. Furthermore, in \cite{alsger2016validating} it is also proven that not only the alighting stops, but also the (last) destination stop of a trip can be estimated from boarding data gathered by a smart card, within a reasonable accuracy.

This summarized review proves that, although we were not able to access real data from a public transportation company for this work, such curated information about users' trips can indeed be obtained and used for our proposed representations.

@@

\subsubsection{Compressed Suffix Array (CSA)}
\label{sec:csa}
One common structure to count or locate all the occurrences of a pattern $P[1,m]$ in a string $S[1,n]$ of symbols from an alphabet $\Sigma$ is the Suffix Array (SA) \cite{MM93}. The SA lexicographically sorts all the suffixes $\{S[i..n] : 1 \leq i \leq n\}$ and builds an array $A[1,n]$ with their indices, holding that $S[A[i]..n] \prec S[A[i+1]..n]$. This allows to find out all the occurrences of a substring $P[1,m]$ in $O(m \log n)$ time with binary searches.

A simple enhancement to avoid storing the original string $S$ is to align $A$ with a bitvector $D[1,n]$ where $D[i]=1~\Leftrightarrow~S[A[i-1]] \prec S[A[i]]$. That is, $D$ marks with a 1 all the indexes $i$ of $A$ where the starting symbol of that suffix $S[A[i]]$ has a larger lexicographical value than the one from the previous suffix $S[A[i-1]]$. With $D$, keeping $S$ is not longer needed as $S[A[i]] = rank_1(D,i)$.
%, where $rank_1(D,i)$ returns the number of set bits in $D[1..i]$.

@@

\subsubsection{Wavelet Matrix (WM)}
\label{sec:wm}
Based on the Wavelet Trees \cite{WT03}, the Wavelet Matrix \cite{CNO15} is a structure that is able to perform, given a sequence $S[1..n]$ over the vocabulary $\Sigma$, operations of $rank_a(S,i)$ and $select_a(S,i)$ $\forall a \in \Sigma$ in $O(log\sigma)$ time while using $n\lceil log_2\sigma\rceil~+~o(n)$ bits of space, begin $\sigma = |\Sigma|$.

The WM is built with $\lceil log_2\sigma\rceil$ levels, each level having a bitvector $B_i$ with rank and select capabilities, where $B_1$ contains the most significant bit of every symbol $a \in S$, $B_2$ contains the second most significant bit and so forth. To further explain how these bits are positioned, we must consider $a_i$ as the $i$th bit of $a$ in significance, and $B^{-1}_i[a_i]$ as the position of the bit $a_i$ from $a$ in the bitvector $B_i$. Then, 
\[
    B_i[j] = \left\{\begin{array}{lr}
        S[j]_1, & \text{iff } i=1\\
        a_i, & \text{iff } i>1 \wedge (a_{i-1}=0 \wedge j=rank_0(B_{i-1},B^{-1}_{i-1}(a_{i-1}))) \\
        & \vee~(a_{i-1}=1~\wedge~j=rank_0(B_{i-1},n)~+~rank_1(B_{i-1},B^{-1}_{i-1}(a_{i-1})))
        \end{array}\right.
\]

@@

This structure allows to answer the following operations in $O(\log\sigma)$ time:
\begin{itemize}
    \item $access(S,i)$ returns the value in any position in the original sequence $S$ by following through the levels of the WM using the $rank$ operation on bitvectors. If $B_1[i]$ is 0, then $B_2[rank_0(B_1,i)]$ must be accessed next. Otherwise, it is $B_2[rank_0(B_1,n) + rank_1(B_1,i)]$. This must be recursively applied until the leaf level is reached, and the traversed bits will compose the value of $S[i]$.
    \item $rank_a(S,i)$ is calculated with the same operations as in $access$, with the difference that in this case we know whether we need to use $rank_0$ or $rank_1$ on each bitvector as the value of the symbol $a$ is known. To calculate the number of occurrences of $a$ in $S[1..i]$ we must simply subtract the position reached in the leaf level to the position of the first occurrence of $a$ on that level, which can be either stored for each symbol in $O(\sigma\log n)$ bits or calculated by making an analogous traversal from the first position of $S$.
    \item $select_a(S,i)$ first $i$ must be added to the first position of $a$ in the leaf level, which as for $rank$, can be either stored or calculated. After obtaining this position $j$, the WM must be traversed in reverse order using $select$ on bitvectors until the first level is reached. When $a_i=0$ the traversal from the lower level position $B_{i+1}[j]$ must be as $select_0(B_i,j)$, while when $a_i=1$, it is done as $select_1(B_i,j-rank_0(B_{i},n))$.
    \item $range_{a,b}(S,i,j)$ returns the number of occurrences of the symbols between $a$ and $b$ in $S[i..j]$. This operation was first introduced in \cite{gagie2012new} for Wavelet Trees and later supported by the Wavelet Matrix \cite{CNO15}, and we make extensive use of it in this work. It is solved with two simultaneous traversals similar to $rank$ over $i$ and $j$, only that it is possible that at some levels it is required to calculate both $rank_0$ and $rank_1$. However, the worst case time complexity of $O(\log\sigma)$ is maintained, due to the fact that it is not necessary to reach the leaf level for all the symbols in $a..b$\footnote{As long as $a\neq b$. Otherwise, $rank_a(S,j)-rank_a(S,i)$ can be used.}.
\end{itemize}

@@

\section{Our model for compact trip representation}
In our previous structures introduced in \cite{brisaboa2018new}, we propose a work-in-progress model to represent the demand information for a public transportation network, as seen in the Figure~\ref{fig:er}.

@@

These are the main elements from our model:
\begin{itemize}
    \item A \textbf{stop\_place} is a physical stop with a location, on which several lines may make stops.
    \item A \textbf{line} is an ordered sequence of stop places that can be traveled by a transport vehicle, such as a bus or a train. It only considers one direction. For this reason, there is often a different and complementary line for the opposite direction, as shown in the Figure~\ref{fig:example_network}.
    \item A \textbf{journey} is a singular traversal of a transport vehicle over a line. It can be seen as a vehicle trip, instead of a user trip.
    \item A \textbf{stage} is formed by a boarding from a stop and alighting to another from the same single line and journey.
    \item An \textbf{user\_trip} is a concatenation of several stages, until the final destination (alighting stop of the last stage) is reached.
\end{itemize}

@@

This approach allows us to treat the information in a layered fashion: the bottom layer is a static network representation, formed by the line and stop\_place entities, the middle layer are the journeys made by vehicles that make stops at specific times, while the top layer are the trips made by the users over these vehicle journeys. In the Figure~\ref{fig:example_network} we show an example for the two bottom layers, where Line 1R is the reversed of Line 1. Our proposed implementation can be found in Section~\ref{sec:ps}.

@@

\section{Proposed Structures}
\label{sec:ps}
As of today, we are not aware of any indexing structure that would allow to efficiently represent trajectories that could also support all the kinds of queries described in Section~\ref{sec:rq}. For this reason, we propose two data structures, \acumm~and~\ttctr. The former is targeted for queries of type A, solving most aggregation queries in constant time, while the later can be used for queries of type B. Finally, we introduce a more versatile alternative to \ttctr~that we call \ctr.

\subsection{Common Structures}
\label{sec:cs}
Both \acumm~and \ttctr~need a network representation to be operated, along with information about the transport vehicle journeys. Therefore, the following structures represent a network with stops $s_i \in S = \{1,...|S|\}$, lines $l_i \in L = \{1,...|L|\}$ and journeys $j_i \in J^l = \{1,...|J^l|\}$. It is important to state that journeys are \textbf{not} identified by $j_i$, as the same $j_i$ can belong to several $J^l$ from different lines, so we speak about journey \textbf{codes} (jcodes) instead of journey identifiers.

@@

\ttctr~was introduced in \cite{brisaboa2018new}, where the spatial component (the pairs $(s,l)$ for the stops and lines of a trip) is represented with a \texttt{CSA} where each unique pair $(s,l)$ is encoded as an integer $id$ in the input sequence $T[1,n]$ that is used to build the \texttt{CSA}.

@@

\begin{figure}[hbt!]
\includegraphics[width=0.4\textwidth]{example_trips.eps}
\caption{An example five trips sorted as required by \ttctr~and \ctr}
\label{fig:example_trips}
\end{figure}

We also need an injective function to encode the pairs $(s,l)$. Consider a vocabulary $V$ such that:
\begin{itemize}
	\item Entry $V[0]$ is reserved for the terminator symbol $\$$.
	\item Entries $\langle V[1],V[2], \dots V[|S|]\rangle]$ are associated to stops $s_1,s_2,\dots, s_{|S|}$ and are used to represent the final stops of the trips. That is, when a given stop $s_i$ ends a user trip, it is given $id \leftarrow s_i$.
	\item The following $|L|$$\times$$|S|$ entries are associated to the sequence composed of the pairs $(s,l) \in S\times L$, sorted first by the stop id $s$ and later by the line id $l$. That is, entry $V[|S|+1]$ is given to $(s_1,l_1)$; $V[|S|+2]$ to $(s_1,l_2)$; $V[|S|+3]$ to $(s_1,l_3)$; $\dots$; $V[|S|+|L|]$ to $(s_1,l_{|L|})$;  $V[|S|+|L|+1]$ to $(s_2, l_1)$, $V[|S|+|L|+2]$ to $(s_2, l_2)$, and so on. Therefore, it is easy to see that any $(s_i,l_j)$ is going to be associated to the entry $V[|S|+ |L|(i-1) + j]$.
\end{itemize}

While this arrangement would theoretically produce many entries in $V$ that are mapped to pairs $(s,l)$ that are unused in $T$, either because the stop is never traversed by that line or because we do not have the record of a user trip containing it, these entries can be skipped with a compact bitvector $B$ with rank and select capabilities, that marks with a one every used entry from $V$. This will enable us to operate with a much smaller vocabulary $V'$ with only the used entries from $V$, such that $V[i] = V'[rank_1(B,i)]$. Refer to the vocabulary shown Figure~\ref{fig:ttctr}(2) for an example where pairs $s:l$ are encoded to 43 unique identifiers in $V$. After that, $B$ marks which of the entries of $V$ actually appear in the original sequence. Finally $V'$ will contain only 12 entries, for each set bit from $B$. Note that neither $V$ nor $V'$ are explicitly represented in practice, as $rank$ and $select$ operations over $B$ are enough to map and unmap, respectively, vocabulary identifiers.

@@

After this, the sequence $T[1,n]$ is built, with the identifiers obtained from mapping mapping to the vocabulary entries of $V'$, over which a \texttt{CSA} is built, as seen in Figure~\ref{fig:ttctr}(3). Each encoded trip in $T$ is terminated with with additional $\$$ symbols. While in the final \texttt{CSA} we assign all these $\$$ a lexicographical value of 0 $(V[0])$, we assign them different values during the construction of the suffix array (A) to ensure that the entries for $\$$ in $A$ maintain the same order as in the original text. Finally, we make a modification on $\Psi$ to make the entries of each $\$$ point to the start of its own trip instead of the next one. These two modifications are proven necessary for our implemented queries, at the expense of losing some of the properties of a classic CSA. For reference, in Figure~\ref{fig:ttctr} we also present $A'$ and $\Psi'$, that show how our modifications compare to the original CSA.

@@

With \ttctr~we obtain a clear separation between the spatial representation of the trips (\texttt{CSA}) and the temporal representation (\texttt{WM} of $jcodes$), where the former can be used to address queries such as ``number of passengers that started their trip from a stop $X\in S$ and a line $l\in L$'' (\texttt{start\_X$_{L}$} from Section~\ref{sec:rq}) with a binary search of the pattern $\$,X_l$, while the later can be used to filter down these results to a time window (\texttt{start\_X$_{LT}$}) with a $range_{a,b}(S,i,j)$ operation over the \texttt{WM}, where $a$ and $b$ are $jcodes$ obtained from Algorithm~\ref{alg:jcodes} and $i$ and $j$ delimit the range of the results obtained in $\Psi$. Because the $\$$ symbols were made cyclical in $\Psi$, it is also possible to answer \texttt{from\_X\_to\_Y} queries by searching for a pattern $Y,\$,X_l$ instead.

\subsection{\ctr}
\label{sec:ctr}
A fundamental weakness of \ttctr~is that it requires several binary search operations over the \texttt{CSA} in the following cases:
\begin{itemize}
    \item We are interested in the number of passengers that started their trip at a stop $s$ and a time window $t_a...t_z$, but from \textbf{any line} (\texttt{start\_X$_{T}$} from Section~\ref{sec:rq}). As $jcodes$ are relative to lines, we must make a separate query for each possible $(X_{l_i}) \forall l_i \in L$ combination.
    \item We need to restrict the line of a final stop, in queries such as \texttt{end\_X$_{L}$} or \texttt{from\_X\_to\_Y$_{L}$} (and similar variations). Because the final stops belong to separate entries of the vocabulary that do not encode line identifiers, to restrict a stop $Y\in S$ to a line $l\in L$ we need to search for every possible expanded pattern $W_l,Y...$, for every stop $W$ from the line $l$ that could have been boarded before alighting at $Y$. While it looks tempting to address this issue by modifying the design of \ttctr~so that final stops also encode line identifiers, this would in turn make queries that do not restrict the line of the final stop inefficient, and we would need to perform a new query for every combination of $(s_{l_i}) \forall l_i \in L$, as in the previous case. \marginpar{\tiny Viendo los experimentos, uno se puede preguntar si no habria quedado mejor meter lineas en paradas finales...}
\end{itemize}

These weaknesses motivated us to develop a second version, \ctr, which reduces the complexity of the queries that restrict the final line, and delegates line checks on a new WM, allowing for more space-time trade-off. As in \ttctr, the input trips need to be sorted by the same criteria, but in \ctr~we use three complementary structures to represent each component of the sequence, as shown in Figure~\ref{fig:example_xctr}:
\begin{enumerate}[(i)]
    \item An adapted Compressed Suffix Array (\texttt{CSA}) over the stop identifiers of all trips, concatenated into a string with additional terminator symbols $\$$ appended at the end of each trip. As in the \texttt{CSA} from \ttctr, we make these $\$$ symbols maintain the order of the trips and cyclical in $\Psi$. Because this time we do not encode line and stop identifiers together and \texttt{CSA} only encodes stops, there is no need for a complex vocabulary anymore.
    \item \texttt{WML}: Aligned to the entries of (i) there is a Wavelet Matrix (WM) for the line identifiers of each stop. Aligned to the $\$$ section we duplicate the starting lines of each trip. As a trivial optimization, we build a separate WM for every stop, allowing us to save space due to the fact that a single stop does not usually belong to many lines, thus the average height of these WM is no larger (and usually smaller) than the height of a single WM.
    \item \texttt{WMJ}: A WM of jcodes aligned to the leaves of (ii). Note that this makes this structure dependant on (ii), which is coherent with the fact that journey codes themselves are relative to the line identifier. In case (ii) implements the optimization described, the entries of the WM must also be rearranged to match the delimited stops.
\end{enumerate}

@@

    \item In \textbf{line 16} we return the number of entries of \texttt{WMJ} between \DataSty{j$_a$} and \DataSty{j$_z$} within our final subrange. The function \FuncSty{GetCount} does not report any range boundaries as \FuncSty{GetRange} does, but simply returns the number of occurrences. Of course, the time complexity of this operation is still proportional to the height of the WM.
    %even though the matching entries can be scattered (not delimiting a contiguous range).
\end{itemize}
